Can we distinguish modes of selective interactions using linkage disequilibrium?

### 2021-03-25 Manuscript.Rmd
edits throughout

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium between selected variants to test for selective interference and
epistatic interactions, with some disagreement over the interpretation of
observations from data. Interpreting this data is hindered by the relative lack
of analytic or even numerical expectations for patterns of variation between
pairs of loci under the combined effects of selection, dominance, epistasis,
and demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

Non-additive effects, including dominance (i.e., interactions *within* loci)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and the mutation load of strongly damaging
disease mutations [@Clark1998-kq]. On the other hand, epistasis differentially
impacts the deleterious load for asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite appreciation
of the effect of dominance on linked variation [@Turelli2000-kz; @Zhao2016-bb]
and the evolutionary importance of epistatic interactions, we currently lack
models for predicting correlation patterns between linked mutations under
general selection models.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
synergistic epistasis, where multiple deleterious mutations had a greater
effect on fitness than the multiplication of the observed effects of the
individual mutations. The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems which can be compared to evolutionary theory
[@Otwinowski2018-zk].

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed mutations. Motivated by theory that linked negatively selected
mutations will display negative LD due to interference [@Hill1966-gv], and that
epistasis will drive expected LD away from zero, a number of recent studies
have used patterns of LD within classes of putatively selected variants to
infer modes of selective interactions. In a notable study from
@Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster more
than expected along lineages in the *Drosophilid* species complex, and that
those clustered mutations tended to preserve the charge of the protein and were
in positive LD compared to pairs involving synonymous mutations, suggesting
that compensatory nonsynonymous variants are tolerated and maintained. More
recently, @Taverner2020-lk replicated this finding across a diverse set of
genera, showing that such epistatic interactions are important for protein
evolution.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their rarity and low allele frequency mean that statistical measurement of LD
and other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and test inference methods, but do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries (such as $\E[D]$) or slices of
the distribution (Figures \@ref(fig:fig1) and \@ref(fig:hudson) ) instead of
the full three-dimensional distribution. Here, I focus on low-order LD
statistics, such as $\E[D]$ and $\E[D^2]$, and their decay with recombination
distance as these are statistics that are commonly used to test for
interactions between loci. Instead of unnormalized $\E[D^2]$ and $\E[D]$, I
consider expectations for $\sigma_d^2 = \E[D^2]/\E[p(1-p)q(1-q)]$ and
$\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$ and $q$ are the derived allele
frequencies at the left ($A/a$) and right ($B/b$) loci, respectively. These
normalized statistics have two benefits: first, the mutation rate cancels so
that expectations are robust to assumptions about the per-base mutation rate,
and second, analytic expectations for these quantities are available under
neutrality at constant population size for comparison [@Ohta1971-yd].

@@

(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
recombination distance, and right panels show expectations for $\sigma_d^1 =
\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
All expectations are at stationarity, and dashed lines are expectations under
neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
and in (B), we see that the negative LD effect of Hill-Robertson interference
is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
fitness model ($\epsilon=0$). Epistasis also pushes signed LD ($\sigma_d^1$)
away from zero, with synergistic epistasis decreasing LD for negatively
selected mutations and antagonistic epistasis increasing LD. The deviation of
$\sigma_d^1$ from zero under epistasis can extend for long recombination
distances, particularly for stronger selection.

@@

The effects of epistasis between variants on LD have been widely studied.
However, the combined effect of non-additive selection within a locus and
interference has received less attention. @Garcia2020-zn used large-scale
forward simulations to explore the effect of dominance on patterns of LD,
showing that LD depends on both the magnitude of the selection and dominance
coefficients in a nonlinear way, but note that expensive large-scale
simulations are needed to explore these effects as no closed form approach to
computing those expectations exist. The method presented here provides such a
closed form approach, so that expectations of signed LD and other two-locus
statistics can be computed more efficiently and accurately.

@@

While dominance and epistasis are well-characterized models of selection, a
large family of selection models can be specified by assigning unique
selection coefficients to each possible diploid pairs of haplotypes. If we assume
the diploid genotype that is homozygous for the ancestral alleles ($ab/ab$) has
fitness 1, then there are nine other possible diploid two-locus genotypes that
could be given unique fitnesses (Table \@ref(tab:selmodels)), noting that
$AB/ab$ and $Ab/aB$ genotypes can have differing fitness effects.

The case with $s_{AB/ab}\not=s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
models can differ. With a highly parameterized space of possible general
diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected signed LD.

@@

The moment system for $\Psi_n$ readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \@ref(fig:fig3)A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide genealogy
reconstruction [@Speidel2019-nj] applied to the
@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \@ref(fig:fig3)B).

@@

From gene-wide averages, LD between pairs of missense mutations does not appear
to be different from pairs of synonymous variants, but LD between
loss-of-function variants is substantially more negative. Hill-Robertson
effects under additive selection are expected to be strongest for slightly to
moderately deleterious variants (with $s\sim 1/N_e$), as strongly deleterious
mutations are not expected to interfere with one another [@McVean2000-ox].
However, loss-of-function mutations do not typically fall within this regime,
and inference of the distribution of fitness effects (DFE) for new
loss-of-function variants shows that a large majority are strongly deleterious
(Supporting Information). Instead, negative synergistic epistasis between
strongly deleterious mutations does produce large negative deviations of mean
LD. Strongly deleterious *recessive* mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for $\E[D]$, especially
for loss-of-function variants for which only a few hundred within-gene pair in
human populations.

@@

There are a number of possible interaction scenarios that can result in
positive signed LD between linked loci. One explanation is a prevalence of
pairs of compensatory mutations that are tolerated to co-segregate at high
frequencies within conserved domains [@Yeang2007-gj; @Ivankov2014-tn].
@Callahan2011-ac and @Taverner2020-lk have proposed such a mechanism to explain
observed clusters of mutations in lineages of *Drosophila* and other species.
Another possibility is a model of antagonistic, or diminishing returns
epistasis, in which a single amino acid-changing mutation within a domain
damages the functionality of that subunit, but additional mutations within that
same domain reduce fitness by a factor less than the first mutation. A third
possibility, related to the antagonistic epistasis case, is that selection acts
on the functional domain as a unit instead of at loci within the domain
individually (such as gene-based dominance). In this scenario, double
heterozygotes have different fitnesses depending on whether the mutations are
found on same haplotype or on different haplotypes.

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but at the same distance as
those within the same domain, missense variants have considerably lower LD than
synonymous variants. This difference between synonymous and missense pairs of
variants largely disappears for SNPs separated by more than a few hundred or a
thousand base pairs (Figure \@ref(fig:lddecay)). This suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed, although the strength of epistasis
is also likely to be a function of distance between mutations.

@@

One proposal is that spatial population structure may be responsible for the
increase of LD observed between synonymous variants, as every natural
population lives in some geography that will necessarily induce some structure.
However, neutral evolution alone cannot create this effect. Under standard
population genetic assumptions of constant mutation and recombination rates and
independent mutation events, migration, admixture, and spatial structure do not
result in nonzero $\E[D]$. Rather, for a set of mutations with *given*
differences in allele frequencies between source populations, $D$ can be
nonzero in an admixed population between those particular mutations even when
$D$ is zero in the source populations [@Cavalli-Sforza1971-jv]. But this result
relies on conditioning on directional differences in allele frequencies between
populations. Expected mean allele frequency difference in the source
populations is zero ($\E[(p_1-p_2)(q_1-q_2)] = 0$), as differences are equally
likely to be positive or negative, and resulting $\E[D]$ is still zero when
taking genome-wide averages. Instead, @Sohail2017-zq and @Sandler2020-vk used
forward simulations under multi-population or explicitly spatial models and a
multiplicative fitness function to explore the joint effect of structure,
assortative mating, and interference, scenarios that can lead to nonzero LD.
@Sandler2020-vk find positive LD among neutral mutations in their simulation,
although it extends over much larger distances than observed in the data.
Instead, positive LD in data between synonymous mutations decays to zero within
just a few hundred bases (see Figures \@ref(fig:lddecay) and
\@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).

@@

The moment system for $\Psi_n$ can be derived directly from the diffusion
approximation, or it can be found through a more intuitive process of tracking
the dynamics of allelic states of a sample of size of $n$ from the full
population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/Ne)$ so that
multiple coalescence, recombination, or selective events within the $n$
lineages are rare in any given generation. For example, for drift events, we
account for the probability that two lineages within the $n$ tracked lineages
at time $t+1$ coalesced in the previous generation and, given a coalescence
event, the probability that haplotype counts changed due to this coalescence
event [see Supporting Information; @Jouganous2017-pq; @Ragsdale2019-nt]. In
typical diffusion approximation fashion, we multiple through by $2N_{ref}$ so
that time is measured in $2N_{ref}$ generations, and we consider scaled
parameters $\rho = 4Nr$, $\theta = 4Nu$, and $\gamma=2Ns$.

@@

In the absence of selection and for fully linked loci (i.e., $\rho=0$), the
system is closed and can be solved exactly. However, for nonzero recombination
or selection, the entries of $\Psi_n$ rely on the slightly sampling
distributions with sample sizes $n+1$ (for recombination and additive
selection) or $n+2$ (for non-additive selection). This is because if a
recombination event occurs within one of $n$ lineages being tracked by
$\Psi_n$, we need to draw an additional lineage from the full population to
recombine with that chosen lineage, thus requiring $\Psi_{n+1}^t$ to find
$\Psi_n^{t+1}$. Selection events similarly require extra lineages from the full
population, which replace a chosen lineage that fails to reproduce with
probability proportional to its selection coefficient, assuming negative
selection.

@@

The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
using nearby entries in $\Psi_n$, is more accurate for larger sample sizes
(Figure \@ref(fig:jackknife)), creating a tension between efficiency and
accuracy: larger sample sizes result in more accurate solutions, as error in
the jackknife is diminished, but computational complexity also grows rapidly in
the number of entries of $\Psi_n$, which is $O(n^3)$. Accuracy and runtime are
shown in Figure \@ref(fig:jackknife). In the results presented in this paper,
sample sizes between $n=30$ and $n=80$ are used.

Derivations for the drift, recombination, and mutation operators and the
jackknife moment-closure approximation can be found in section S1.3 of
@Ragsdale2019-nt, and I repeat the main results in the Supporting Information
of this paper.

@@

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, which has an expected sampling distribution
of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the relative fitness of its haplotype state. If a lineage is
rejected, a replacement is drawn from the full population. Under the assumption
that $ns_{\cdot\cdot} \ll 1$, the probability that more than one selection
event occurs in any given generation is negligibly small, so that the case of
multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
operator $\mathcal{S}$ for additive selection is given in the Supporting
Information.

To account for dominance within loci, or other general forms of two-locus
selection, the selection operator no longer reduces to individual haplotypes,
but instead we need to know the state of two-locus genotypes. For example, the
fitness of an individual carrying an $Ab$ haplotype depends on whether their
second haplotype is $ab$, $Ab$, $aB$, or $AB$. We can therefore assign a
selection coefficient to each possible diploid configuration, $s_{AB/AB}$,
$s_{AB/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
genotype has relative fitness 1, this gives nine possible unique selection
coefficients in the most general two-locus selection model. Note that $AB/ab$
and $Ab/aB$ genotypes need not have the same selection coefficient, which
allows for simulation under a gene-based model of dominance (Table
\@ref(tab:selmodels)).

@@

I am grateful to Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal
Sohail, Matthias Steinrücken, and Kevin Thornton for helpful discussions and
feedback on earlier versions of this manuscript that improved this work.

>>>>

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and correlations between pairs
of mutations. A number of recent studies have used patterns of linkage
disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by the relative
lack of analytic or even numerical expectations for patterns of variation
between pairs of loci under the combined effects of selection, dominance,
epistasis, and demography. Here, I develop a numerical approach to compute the
expected two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations.
Finally, I use whole-genome sequencing data from humans to assess how well we
can differentiate modes of selective interactions in practice. I find that
positive LD between missense mutations within genes is driven by strong
positive allele-frequency correlations between pairs of mutations that fall
within the same conserved domain, pointing to compensatory mutations or
antagonistic epistasis as the prevailing mode of interaction within but not
outside of conserved genic elements. The heterogeneous landscape of both
mutational fitness effects and selective interactions within protein-coding
genes calls for more refined inferences of the joint distribution of fitness
and interactive effects, and the methods presented here should prove useful in
that pursuit.

@@

Non-additive effects, including dominance (i.e., interactions *within* a locus)
and epistasis (interactions *between* loci), further complicate our
evolutionary models. A large fraction of nonsynonymous coding mutations are
thought to be at least partially recessive [@Agrawal2011-wb; @Huber2018-cp],
with average levels of dominance correlating with strength of selection
[@Kacser1981-nc], and dominance plays an important role in shaping expected
equilibrium allele frequencies and the mutation load of strongly damaging
disease mutations [@Clark1998-kq]. On the other hand, epistasis differentially
impacts the deleterious load in asexually and sexually reproducing organisms
[@Kimura1966-gc; @Kondrashov1995-va], has been invoked as an explanation for
the evolutionary advantage of sex [@Kondrashov1982-sf; @Charlesworth1990-kw;
@Barton1998-lu], and can drive incompatibilities that lead to postzygotic
isolation during the process of speciation [@Turelli2000-kz]. Within
populations, epistasis is known to cause signed LD to deviate dramatically from
zero [@Charlesworth1990-kw; @Kondrashov1995-va]. However, despite appreciation
of the effect of dominance on linked variation [@Turelli2000-kz; @Zhao2016-bb]
and the evolutionary importance of epistatic interactions, we currently lack
models for predicting correlation patterns between linked mutations under
general selection models.

@@

The most direct way to test for interactions between linked selected variants
is through deep mutation scanning experiments, in which many distinct mutations
are introduced within a target gene and then organismal fitness or some protein
function is experimentally measured [@Romero2009-yi; @Bank2015-vq;
@Puchta2016-rx; @Steinberg2016-is]. For example, using the model system of the
TEM-1 $\beta$-lactamase gene in *E. coli*, @Bershtein2006-bk found evidence for
synergistic epistasis, where multiple deleterious mutations had a greater
effect on fitness than the multiplication of the observed effects of the
individual mutations. The scale of mutation scanning experiments continues to
improve dramatically, promising greater resolution of the fitness landscape in
such model systems that can be compared to evolutionary theory
[@Otwinowski2018-zk].

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theory that linked
negatively selected mutations will display negative LD due to interference
[@Hill1966-gv], and that epistasis will drive expected LD away from zero, a
number of recent studies have used patterns of LD within classes of putatively
selected variants to infer modes of selective interactions. In a notable study
from @Callahan2011-ac, pairs of nonsynonymous mutations were found to cluster
more than expected along lineages in the *Drosophilid* species complex, and
that those clustered mutations tended to preserve the charge of the protein and
were in positive LD compared to pairs involving synonymous mutations,
suggesting that compensatory nonsynonymous variants are tolerated and
maintained. More recently, @Taverner2020-lk replicated this finding across a
diverse set of genera, showing that such epistatic interactions are important
for protein evolution.

@@

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their rarity and low allele frequency mean that statistical measurement of LD
and other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and be used to test inference methods, but
they do not efficiently provide expectations for quantities of interest across
the range of relevant parameters. Analytical and numerical methods for expected
haplotype frequencies and LD under general selective interaction models are
thus crucial for interpreting patterns of variation observed in data.

@@

In the Methods, I expand on the moment system developed in @Ragsdale2019-nt to
compute the expected sampling distribution $\Psi_n$ of two-locus haplotypes
under a general model of selective interactions. This sampling distribution
stores the expected density or observed counts of pairs of biallelic loci with
each possible haplotype frequency configuration in a sample of size $n$. In
many cases it is simpler to visualize summaries such as the expectation or
variance of $D$ or slices of the distribution (e.g., Figures \@ref(fig:fig1),
\@ref(fig:fig2) and \@ref(fig:hudson)) instead of the full three-dimensional
distribution. Here I focus on low-order LD statistics including $\E[D]$ and
$\E[D^2]$ and their decay with recombination distance, as these are statistics
that are commonly used to test for interactions between loci. Instead of
unnormalized $\E[D^2]$ and $\E[D]$ I consider expectations for $\sigma_d^2 =
\E[D^2]/\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D]/\E[p(1-p)q(1-q)]$, where $p$
and $q$ are the derived allele frequencies at the left ($A/a$) and right
($B/b$) loci, respectively. These normalized statistics have two benefits:
first, the mutation rate cancels so that expectations are robust to assumptions
about the per-base mutation rate, and second, analytic expectations for these
quantities are available under neutrality at constant population size for
comparison [@Ohta1971-yd].

@@

(ref:fig1cap) \textbf{Additive selection and epistasis.} Left panels show
expectations for the decay of $\sigma_d^2 = \E[D^2] / \E[p(1-p)q(1-q)]$ with
recombination distance, and right panels show expectations for $\sigma_d^1 =
\E[D] / \E[p(1-p)q(1-q)]$. The scaled recombination rate is given by
$\rho=4N_er$, where $r$ is the probability of recombination between two loci.
All expectations are at stationarity, and dashed lines are expectations under
neutrality (Equation \@ref(eq:ohta)). (A-B) Additive selection, with equal
selection strength at both loci. Stronger selection reduces $\sigma_d^2$ (A),
and in (B), we see that the negative LD effect of Hill-Robertson interference
is strongest for intermediate selection ($\gamma \approx -1$ to $-5$, where
$\gamma=4N_es$). (C-F) Epistasis with additive selection. Synergistic epistasis
(\(\epsilon < 0\)) decreases $\sigma_d^2$ while antagonistic epistasis
(\(\epsilon > 0\)) increases it compared to expectations under a multiplicative
fitness model ($\epsilon=0$). (See Methods for the definition of the
epistasis model.) Epistasis also pushes signed LD ($\sigma_d^1$) away from
zero, with synergistic epistasis decreasing LD for negatively selected
mutations and antagonistic epistasis increasing LD. The deviation of
$\sigma_d^1$ from zero under epistasis can extend for long recombination
distances, particularly for stronger selection.

@@

The effects of epistasis on LD have been widely studied. However, the combined
effect of non-additive selection within a locus and interference has received
less attention. @Garcia2020-zn used large-scale forward simulations to explore
the effect of dominance on patterns of LD, showing that LD depends on both the
magnitude of the selection and dominance coefficients in a nonlinear way, but
note that expensive large-scale simulations are needed to explore these effects
as no closed form approach to computing those expectations exist. The method
presented here provides such a closed form approach, so that expectations of
signed LD and other two-locus statistics can be computed more efficiently and
accurately.

@@

Beyond our standard models of epistasis and dominance, a large family of
selection models can be specified by assigning unique fitness effects to each
possible diploid pair of haplotypes. If we assume the diploid genotype that is
homozygous for the ancestral alleles ($ab/ab$) has fitness 1, then there are
nine other possible diploid two-locus genotypes that could be given unique
fitnesses (Table \@ref(tab:selmodels)), noting that $AB/ab$ and $Ab/aB$
genotypes can have differing selection coefficients.

The case with $s_{AB/ab} \not= s_{Ab/aB}$ can arise in a scenario where a
mutation at either locus within a haplotype impacts some functional region or
element, but a diploid individual carrying at least one copy that is free of
mutations has minimal fitness loss. In this "gene-based dominance" scenario
[e.g., @Sanjak2017-zw], an $AB/ab$ genotype has higher fitness than an $Ab/aB$
type (a simple implementation of this model is given in the final column in
Table \@ref(tab:selmodels)). Such a gene-based fitness model gives similar
expected positive signed LD to the model of antagonistic epistasis (compare
Figure \@ref(fig:fig3)E-F to G-H), although the interpretation of those two
models can differ. With a highly parameterized space of possible general
diploid selection models, multiple models with different biological
interpretations can give similar patterns of expected signed LD.

@@

The moment system for $\Psi_n$ readily incorporates variable population size as
well as mutation and recombination rates and selection parameters that change
over time. Here I focus on non-equilibrium population size history and consider
scenarios relevant to human demographic history, including bottlenecks and
expansions. I explore two simple models (Figure \@ref(fig:fig3)A), one with an
instantaneous expansion and another with a bottleneck followed by recovery. I
also consider two demographic histories inferred using genome-wide gene
genealogy reconstruction [@Speidel2019-nj] applied to the
@1000_Genomes_Project_Consortium2015-zq dataset, and focus on size histories
for the Yoruba from Ibidan, Nigeria (YRI) and Utahns of North and West European
ancestry (CEU) (Figure \@ref(fig:fig3)B).

@@

From gene-wide averages, LD between pairs of missense mutations does not appear
to be different from pairs of synonymous variants, but LD between
loss-of-function variants is more negative. Hill-Robertson effects under
additive selection are expected to be strongest for slightly to moderately
deleterious variants (with $s\sim 1/N_e$), as strongly deleterious mutations
are not expected to interfere with one another [@McVean2000-ox]. However,
loss-of-function mutations do not typically fall within this regime, and
inference of the distribution of fitness effects (DFE) for new loss-of-function
variants shows that a large majority are strongly deleterious (Supporting
Information). Instead, negative synergistic epistasis between strongly
deleterious mutations does produce large negative deviations of mean LD.
Strongly deleterious *recessive* mutations can also produce this pattern,
making it difficult to distinguish these effects with certainty. This
difficulty is compounded by the large measurement noise for $\E[D]$, especially
for loss-of-function variants for which only a few hundred within-gene pairs
exist in human population data considered.

@@

There are a number of possible interaction scenarios that can result in
positive signed LD between linked loci. One explanation is a prevalence of
pairs of compensatory mutations that are tolerated to co-segregate at high
frequencies within conserved domains [@Yeang2007-gj; @Ivankov2014-tn].
@Callahan2011-ac and @Taverner2020-lk have proposed such a mechanism to explain
observed clusters of mutations in lineages of *Drosophila* and other species.
Another possibility is a model of antagonistic, or diminishing returns
epistasis, in which a single amino acid-changing mutation within a domain
damages the functionality of that subunit, but additional mutations within that
same domain reduce fitness by a factor less than the first mutation. A third
possibility, related to the antagonistic epistasis case, is that selection acts
on the functional domain as a unit instead of at loci within the domain
individually (such as under a model of gene-based dominance). In this scenario,
double heterozygotes have different fitnesses depending on whether the
mutations are found on same haplotype or on different haplotypes.

We do not see an increase in LD between mutations that are found in different
annotated domains within the same gene, which are on average at much larger
recombination distances. Nor do we find increased LD outside of domains.
Rather, for pairs of mutations outside of domains but matched to the same
distance as those within the same domain, missense variants have considerably
lower LD than synonymous variants. This difference between synonymous and
missense pairs of variants largely disappears for SNPs separated by more than a
few hundred base pairs (Figure \@ref(fig:lddecay)). This suggests that
Hill-Robertson interference is the primary mode of interaction between
mutations falling outside of domains, as epistasis is expected to impact LD
over larger distances than what is observed, although the strength of epistasis
is also likely to be a function of distance between mutations.

@@

One proposal is that spatial population structure may be responsible for the
increase of LD observed between synonymous variants, as every natural
population lives in some geography that will necessarily induce some structure.
However, neutral evolution in structured populations alone cannot create this
effect. Under standard population genetic assumptions of constant mutation and
recombination rates and independent mutation events, migration, admixture, and
spatial structure do not result in nonzero $\E[D]$. Rather, for a set of
mutations with *given* differences in allele frequencies between source
populations, $D$ can be nonzero in an admixed population between those
particular mutations even when $D$ is zero in the source populations
[@Cavalli-Sforza1971-jv]. But this result relies on conditioning on directional
differences in allele frequencies between populations. Expected mean allele
frequency difference in the source populations is zero ($\E[(p_1-p_2)(q_1-q_2)]
= 0$), as differences are equally likely to be positive or negative, and
resulting $\E[D]$ is still zero when taking genome-wide averages. Instead,
@Sohail2017-zq and @Sandler2020-vk used forward simulations under
multi-population or explicitly spatial models and a multiplicative fitness
function to explore the joint effect of structure, assortative mating, and
interference, scenarios that can lead to nonzero LD. @Sandler2020-vk find
positive LD among neutral mutations in a simulation with admixture and linked
selected mutations, although it extends over much larger distances than
observed in the data. Instead, positive LD in data between synonymous mutations
decays to zero within just a few hundred bases (see Figures \@ref(fig:lddecay)
and \@ref(fig:lddecaydomain), and Figure 5 from @Sandler2020-vk).

@@

The moment system for $\Psi_n$ can be derived directly from the diffusion
approximation, or it can be found through a more intuitive process of tracking
the dynamics of allelic states of a sample of size of $n$ from the full
population. We assume $n \ll N_e$, and $r$ and $s$ are $O(1/N_e)$ so that
multiple coalescence, recombination, or selective events within the $n$
lineages are rare in any given generation. For example, for drift events, we
account for the probability that two lineages within the $n$ tracked lineages
at time $t+1$ coalesced in the previous generation and, given a coalescence
event, the probability that haplotype counts changed due to this coalescence
event [see Supporting Information; @Jouganous2017-pq; @Ragsdale2019-nt]. In
typical diffusion approximation fashion, we multiple through by $2N_{ref}$ so
that time is measured in $2N_{ref}$ generations, and we consider scaled
parameters $\rho = 4Nr$, $\theta = 4Nu$, and $\gamma=2Ns$.

@@

In the absence of selection and for fully linked loci (i.e., $\rho=0$), the
system is closed and can be solved exactly. However, for nonzero recombination
or selection, the entries of $\Psi_n$ rely on the slightly larger sampling
distributions with sample sizes $n+1$ (for recombination and additive
selection) or $n+2$ (for non-additive selection). This is because if a
recombination event occurs within one of $n$ lineages being tracked by
$\Psi_n$, we need to draw an additional lineage from the full population to
recombine with that chosen lineage, thus requiring $\Psi_{n+1}^t$ to find
$\Psi_n^{t+1}$. Selection events similarly require extra lineages from the full
population, which replace a chosen lineage that fails to reproduce with
probability proportional to its selection coefficient, assuming negative
selection.

@@

The jackknife approximation, which approximates an entry $\Psi_{n+l}(i,j,k)$
using nearby entries in $\Psi_n$, is more accurate for larger sample sizes,
creating a tension between efficiency and accuracy: larger sample sizes result
in more accurate solutions, as error in the jackknife is diminished, but
computational complexity also grows rapidly in the number of entries of
$\Psi_n$, which is $O(n^3)$. Accuracy and runtime are shown in Figure
\@ref(fig:jackknife). In the results presented in this paper, sample sizes
between $n=30$ and $n=80$ are used. Derivations for the drift, recombination,
and mutation operators and the jackknife moment-closure approximation can be
found in section S1.3 of @Ragsdale2019-nt, and I repeat the main results in the
Supporting Information of this paper.

@@

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, which has an expected sampling distribution of
haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
derived haplotype, each of those sampled lineages has probability of being
rejected equal to the selection coefficient assigned to its haplotype state. If
a lineage is rejected, a replacement is drawn from the full population. Under
the assumption that $ns_{\cdot\cdot} \ll 1$, the probability that more than one
selection event occurs in any given generation is negligibly small, so that the
case of multiple simultaneous rejections can be ignored. Then $\Psi_n^{t+1}$
relies only on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full
selection operator $\mathcal{S}$ for additive selection is given in the
Supporting Information.

To account for dominance within loci, or other general forms of two-locus
selection, the selection operator no longer reduces to individual haplotypes,
but instead we need to know the state of two-locus genotypes. For example, the
fitness of an individual carrying an $Ab$ haplotype depends on whether their
second haplotype is $ab$, $Ab$, $aB$, or $AB$. We can therefore assign a
selection coefficient to each possible diploid configuration, $s_{Ab/ab}$,
$s_{Ab/Ab}$, and so on. Assuming that the doubly homozygous ancestral $ab/ab$
genotype has relative fitness 1, this gives nine possible unique selection
coefficients in the most general two-locus selection model. Note that $AB/ab$
and $Ab/aB$ genotypes need not have the same selection coefficient, which
allows for simulation under a gene-based model of dominance (Table
\@ref(tab:selmodels)).

@@

I thank Alex Diaz-Papkovich, Eric Friedlander, Simon Gravel, Mashaal Sohail,
Matthias Steinrücken, and Kevin Thornton for helpful discussions and feedback
on earlier versions of this manuscript.
